Classification Techniques for Conformance and Performance Checking in Process Analysis

### 2013-12-02 paper/Paper_aibp/camera-ready-paper2.tex
aggiunta versione finale paper e presentazione

After the presentation of a case study in Section~\ref{example}, Section~\ref{Background} introduces some background concepts about process analysis. First the log replay algorithm is briefly described, then conformance  analysis is introduced using the business process of Section~\ref{example}. In Section~\ref{ClassifApproach} our approach to process analysis based on classification is presented. First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study.  A possible extension of the approach for performance analysis is briefly discussed presented. Finally, Section~\ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins, as well the methodology to follow in order to analyse processes using our approach.

\section{The Case Study: a Sale Business Process}\label{example}
A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal. There are several formalisms for representing business processes. In this paper we represent them using Petri nets~\cite{DBLP:conf/ifip/Petri62,DBLP:journals/jcsc/Aalst98} since this makes possible to exploit some existing process analysis techniques that are based on such nets. However, in a business management context, Petri nets can be not expressive enough and usually process models are presented as workflows, for example using a standard notation for the business process modeling: BPMN (Business Process Model and Notation)~\cite{BPMN}. Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN.pdf}
\caption{Petri net for a sale process}
\label{pnet}
\end{figure}
%
Figure~\ref{pnet} presents the Petri net model for a sale
process.\footnote{\small{
The Petri net of Figure~\ref{pnet} is obtained by
transforming a BPMN model using the algorithm of~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
each activity to a pair of start/end transitions. The resulting net
has been simplified by removing the unnecessary \emph{invisible
  transitions} used to encode the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
}} 
This business process abstractly represents a fragment of the procedure followed in a commercial organization for managing orders received from clients. In general this procedure includes several activities and each of them involves a specific department of the organization. We assume that the sale process begins with the notification of an order from a client, so the first activity is simply called \emph{Order}. After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependency. The \emph{FinancialCheck} activity represents the financial analysis of the order (i.e verifying the financial situation of the client). Concurrently, the \emph{WarehouseCheck} activity checks the availability of merchandise requested by the order and, in case of a shortage, starts a supplying procedure. This is done through the activity \emph{ExternalWarehouse}. After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure with the final activity called \emph{Notify}, by which a result regarding the acceptance of the order is communicated to the client.

@@

The general approach for solving a classification problem~\cite{tan2005introduction} consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This step allows to  estimate the \emph{model accuracy} using one of the standard approaches. For example, the accuracy can be represented by the percentage  of records correctly predicted  with respect to the total number of records in the test set.

@@

\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, relevant information is  extracted and included as a record in the dataset. For the process presented in Section~\ref{example}, a record includes  the following attributes: an instance identifier, a  client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. Furthemore, as target attribute we associate with each record the conformance result for that instance,  which is a boolean value that is extracted from the output of the log replay. All these attributes are discrete and contribute to build the dataset
presented in Table~\ref{tab:SaleData} of the classification problem. In this preliminary work,  in order to 
explore the applicability of existing machine learning  techniques to process analysis, we just used synthetic data
generated to emulate process executions. Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities. In general, an attribute can characterize several aspects of an activity.

@@

\section{Implementation with ProM}\label{implementation}
In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6}~\cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka}~\cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
\begin{itemize}
\item \texttt{Generate Instances with Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay. It returns as output a training set as shown in Table~\ref{tab:SaleData} in the format needed by the Weka classification tool. Note that the plug-in extracts \emph{all} the attributes data recorded in the event log. If necessary, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.

\item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
\end{itemize}
The second category includes plug-ins for the classifier generation and utilization as a predictive model:
\begin{itemize}
\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using the $J48$ algorithm, an implementation of the algorithm $C4.5$ provided by Weka.
\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classifies the instances according to the classifier.
\end{itemize}
Finally, the last category includes a set of plug-ins useful for the serialization, deserialization and visualization of the resources used by the previous plug-ins, such as dataset and classifiers.

Figure \ref{dataflow} represents a data flow diagram that summarizes data dependencies between plug-ins discussed in this section.

\begin{figure}[h]
\includegraphics[width=360pt]
{./items/dataflow.pdf}
\caption{Data flow diagram}
\label{dataflow}
\end{figure}

@@

We presented some preliminary results of a research activity aimed at applying machine learning techniques, i.e.~classification algorithms, in the realm of process analysis. The idea is to try to identify some correlation between the data value of certain variables in the process instances of event logs, and the results of conformance checking. A successful identification of such values could allow to predict the conformance result  of an instance during execution.  The same technique was applied to performance checking as well, but only briefly summarized here because of space constraints. The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine Weka. Some guidelines about their use are presented in a concluding section. 
The approach has been successfully tested on synthetic event logs only. An obvious continuation of this work includes experimentations with real event logs: this will allow us to evaluate the scalability of the approach and its robustness in presence of noise.

>>>>

After the presentation of a case study in Section~\ref{example}, Section~\ref{Background} introduces some background concepts about process analysis. First the log replay algorithm is briefly described, then conformance  analysis is introduced using the business process of Section~\ref{example}. In Section~\ref{ClassifApproach} our approach to process analysis based on classification is presented. First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study.  A possible extension of the approach for performance analysis is briefly discussed. Finally, Section~\ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins.
\section{The Case Study: a Sale Business Process}\label{example}
A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal. There are several formalisms for representing business processes. In this paper we represent them using Petri nets~\cite{DBLP:conf/ifip/Petri62,DBLP:journals/jcsc/Aalst98} since this makes possible to exploit some existing process analysis techniques that are based on such nets. However, in a business management context, Petri nets can be not expressive enough and usually process models are presented as workflows, for example using a standard notation for the business process modeling: BPMN (Business Process Model and Notation)~\cite{BPMN}. Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN.pdf}
\caption{Petri net for a sale process}
\label{pnet}
\end{figure}
%
Figure~\ref{pnet} presents the Petri net model for a sale
process.\footnote{\small{
The Petri net of Figure~\ref{pnet} is obtained by
transforming a BPMN model using the algorithm of~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
each activity to a pair of start/end transitions. The resulting net
has been simplified by removing the unnecessary \emph{invisible
  transitions} used to encode the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
}} 
This business process abstractly represents a fragment of the procedure followed in a commercial organization for managing orders received from clients. In general this procedure includes several activities and each of them involves a specific department of the organization. We assume that the sale process begins with the notification of an order from a client, so the first activity is simply called \emph{Order}. After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependency. The \emph{FinancialCheck} activity represents the financial analysis of the order (i.e verifying the financial situation of the client). Concurrently, the \emph{WarehouseCheck} activity checks the availability of merchandise requested by the order and, in case of a shortage, starts a supplying procedure. This is done through the activity \emph{ExternalWarehouse}. After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure with the final activity called \emph{Notify}, by which a result regarding the acceptance of the order is communicated to the client.

@@

The general approach for solving a classification problem consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This step allows to  estimate the \emph{model accuracy} using one of the standard approaches. For example, the accuracy can be represented by the percentage  of records correctly predicted  with respect to the total number of records in the test set.

@@

\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, relevant information is  extracted and included as a record in the dataset. For the process presented in Section~\ref{example}, a record includes  the following attributes: an instance identifier, a  client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. Furthemore, as target attribute we associate with each record the conformance result for that instance,  which is a boolean value that is extracted from the output of the log replay. All these attributes are discrete and contribute to build the dataset
presented in Table~\ref{tab:SaleData} of the classification problem. In this preliminary work,  in order to 
explore the applicability of existing machine learning  techniques to process analysis, we just used synthetic data
generated to emulate process executions. Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities, but in general attributes can characterize several other aspects of activities.

@@

\section{Implementation with ProM}\label{implementation}
In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6}~\cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka}~\cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
We can classify the developed plug-ins in three categories. The first one includes plug-ins for building the dataset needed for the classification:
\begin{itemize}
\item \texttt{Generate Instances with Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay. It returns as output a training set as shown in Table~\ref{tab:SaleData} in the format needed by the Weka classification tool. Note that the plug-in extracts \emph{all} the attributes data recorded in the event log. If necessary, the resulting dataset can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.

\item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
\end{itemize}
The second category includes plug-ins for generating the classifier and for using it as a predictive model:
\begin{itemize}
\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using the $J48$ algorithm, an implementation of the algorithm $C4.5$ provided by Weka.
\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classifies the instances according to the classifier.
\end{itemize}
Finally, the last category includes a set of plug-ins useful for the serialization, deserialization and visualization of the resources used by the previous plug-ins, such as dataset and classifiers.

Figure \ref{dataflow} represents a data flow diagram that summarizes the data dependencies among the plug-ins discussed in this section.

\begin{figure}[h]
\includegraphics[width=360pt]
{./items/dataflow.pdf}
\caption{Data dependencies among the implemented plug-ins.}
\label{dataflow}
\end{figure}

@@

We presented some preliminary results of a research activity aimed at applying machine learning techniques, i.e.~classification algorithms, in the realm of process analysis. The idea is to try to identify some correlation between the data value of certain variables in the process instances of event logs, and the results of conformance checking. A successful identification of such values could allow to predict the conformance result  of an instance during execution.  The same technique was applied to performance checking as well, but only briefly summarized here because of space constraints. The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine Weka. The approach has been successfully tested on synthetic event logs only. An obvious continuation of this work includes experimentations with real event logs: this will allow us to evaluate the scalability of the approach and its robustness in presence of noise.
